---
title: "Covid-19 and Election Data"
output: flexdashboard::flex_dashboard
---
Map {data-orientation=rows}
=====================================   

```{r setup, include = FALSE}
library(tidyverse)
library(lubridate)
library(datasets)
library(tidycensus)
library(dygraphs)
library(xts)
library(ggbeeswarm)

#Initial reading of the main data file
covid_counties <- read_csv("data_raw/us-counties.csv")
election_counties <- read_csv("data_raw/presidential_election_counties.csv")
masks_counties <- read_csv("data_raw/mask-use-by-county.csv")
# This is from :
#https://github.com/balsama/us_counties_data/blob/main/data/counties.csv#L15
population_counties <- get_estimates( geography = "county", year = "2019", variables = "POP")

#New York population data was incorrect from above source
#State population data pulled from census API
#install census api key for future session use
#census_api_key(key = "..........", install = TRUE)
#run this to load census api key and allow queries of census data
readRenviron("~/.Renviron")

#getting state population for 2019
state_population <- get_estimates( geography = "state", year = "2019", variables = "POP")

#data from fips_code and joining it to population_counties 
fips_codes <- tibble(fips_codes)

#data from satasets package to add 2 letter state abbr. column
states = tibble(state_abb = state.abb, state = str_to_lower(state.name)) %>%
  mutate(state_abb = factor(state_abb))


#-------------------------------------------------------------------------------

#creating 5 digit fips code and join to population_counties 
#16 counties in fips_code and not in population_counties
fips_codes_clean <- fips_codes %>%
  rename_all(str_to_lower) %>% 
  select(state_name, county, state_code, county_code) %>% 
  mutate(fips = paste0(state_code, county_code), 
         county =str_to_lower(county), 
         state_name  = str_to_lower(state_name),
         state = state_name) %>% 
  select(fips, state, county)

#rename specific values
fips_codes_clean[fips_codes_clean == "petersburg census area"] <- "petersburg borough"

fips_codes_clean[fips_codes_clean == "la salle parish"] <- "lasalle parish"


#Cleaning up county population data
#Join for final population_counties_clean

population_counties_clean <- population_counties %>% 
  separate(NAME, c("county", "state"), sep = ", ") %>% 
  rename_all(str_to_lower) %>% 
  select(-geoid, -variable) %>% 
  mutate(county = str_to_lower(county),
         state = str_to_lower(state),
         population = value) 

#cleaning accents
population_counties_clean[population_counties_clean == "doña ana county"] <- "dona ana county"

population_counties_clean <- population_counties_clean %>% 
  left_join(fips_codes_clean, by = c ("county", "state")) %>% 
  select (-value, -county, -state)

#Selecting only Trump & Biden results (along with other county identifying columns)
#Also creating percent_biden column that can be used for gradient stuff
election_counties_clean <- election_counties %>% 
  select(fips, name, state, results_trumpd, results_absentee_trumpd,
         results_bidenj, results_absentee_bidenj) %>% 
  mutate(name = str_to_lower(name),
         trump_total = results_trumpd + results_absentee_trumpd,
         biden_total = results_bidenj + results_absentee_bidenj,
         percent_biden = biden_total / (biden_total + trump_total),
         winner = if_else(biden_total > trump_total, "biden", "trump"),
         winner = factor(winner),
         state = str_replace(state, "-", " ")) %>% 
  rename(county = name) %>%
  filter(state != "district of-columbia")


### Retaining date column and changing date column to Date class
covid_counties_clean <- covid_counties %>% 
  filter(date == "2020-11-17") %>% 
  select(fips, county, state, cases, deaths) %>% 
  mutate(county = str_to_lower(county),
         state = str_to_lower(state))

covid_counties_clean_dates <- covid_counties %>% 
  mutate(date = ymd(date),
         county = str_to_lower(county),
         state = str_to_lower(state))


# CHECKING TO MAKE SURE CUMULATIVE AND NOT ACTIVE CASES
# covid_counties %>% 
#   filter(fips == "04013") %>% 
#   ggplot(aes(date, cases)) +
#   geom_lin

# each participant was asked: How often do you wear a mask in public when you expect to be within six feet of another person?
# 250,000 survey responses between July 2 and July 14
# mask_percent is just what I thought would give the % of people who "commonly" wear a mask in the county when in public
masks_counties_clean <- masks_counties %>% 
  rename_all(str_to_lower) %>% 
  rename(fips = countyfp) %>% 
  mutate(mask_percent = frequently + always)

#Combining Data Sets (ALASKA does not have accurate county population information)
#We have 40 counties in Alaska with election data, but no county population data 
master_covid_election <- covid_counties_clean %>% 
  inner_join(population_counties_clean, by = "fips") %>% 
  left_join(election_counties_clean, by = c("fips", "county", "state")) %>% 
  inner_join(masks_counties_clean, by = "fips") %>%
  inner_join(states, by = "state")

master_covid_election_with_dates <- covid_counties_clean_dates %>% 
  inner_join(population_counties_clean, c("fips")) %>% 
  left_join(election_counties_clean, by = c("fips", "county", "state")) %>% 
  inner_join(masks_counties_clean, by = "fips") %>%
  inner_join(states, by = "state")

sum(is.na(master_covid_election_with_dates$fips))

# #Just looking at what counties had the highest death rate
# master_covid_election %>% 
#   mutate(death_rate = (deaths / population) * 100) %>% 
#   select(state, county, death_rate) %>% 
#   arrange(desc(death_rate))


#states won biden or trump. source = https://www.270towin.com/2020-election-results-live/
biden_states <- c("WA", "OR", "CA", "NV", "AZ", "NM", "CO", "HI", "MN", "WI", 
                  "IL", "MI", "GA", "PA", "VA", "NY", "VT", "NH", "MA", "RI", 
                  "CT", "NJ", "DE", "MD", "DC")
trump_states <- c("AK", "ID", "UT", "MT", "WY", "ND", "SD", "KS", "OK", "TX", 
                  "LA", "IA", "MO", "AR", "MS", "AL", "TN", "KY", "IN", "OH", 
                  "WV", "NC", "SC", "FL")
split_states <- c("NE", "ME")

#adding state win column
master_covid_election <- master_covid_election %>%
  mutate(state_win = case_when(state_abb %in% biden_states ~ "biden",
                               state_abb %in% trump_states ~ "trump",
                               state_abb %in% split_states ~ "split",
                               is.na(winner) & state %in% biden_states ~ "biden",
                               is.na(winner)& state %in% trump_states ~ "trump",
                               is.na(winner)& state %in% split_states ~ "split"))

master_covid_election_with_dates <- master_covid_election_with_dates %>%
  mutate(state_win = case_when(state_abb %in% biden_states ~ "biden",
                               state_abb %in% trump_states ~ "trump",
                               state_abb %in% split_states ~ "split",
                               is.na(winner) & state %in% biden_states ~ "biden",
                               is.na(winner)& state %in% trump_states ~ "trump",
                               is.na(winner)& state %in% split_states ~ "split"))

#adding total cases per state column
total_state_cases <- master_covid_election %>%
  group_by(state_abb) %>%
  summarise(total_cases = sum(cases))

master_covid_election <- master_covid_election %>%
  inner_join(total_state_cases, by = "state_abb") %>%
  mutate(state = factor(state),
         state_abb = factor(state_abb))

total_state_cases <- master_covid_election_with_dates %>%
  group_by(state_abb, date) %>%
  summarise(total_cases = sum(cases))

master_covid_election_with_dates <- master_covid_election_with_dates %>%
  inner_join(total_state_cases, by = c("state_abb", "date")) %>%
  mutate(state = factor(state),
         state_abb = factor(state_abb))

# replacing NAs in winner column with info from state_win
master_covid_election <- master_covid_election %>%
  mutate(winner = case_when(is.na(winner) ~ state_win,
                            !is.na(winner) ~ as.character(winner)),
         death_rate_per_100k = (deaths / population) * 100000,
         infection_rate_per_100k = (cases / population) * 100000,
         death_rate = (deaths / population),
         infection_rate = (cases / population))

master_covid_election_with_dates <- master_covid_election_with_dates %>%
  mutate(winner = case_when(is.na(winner) ~ state_win,
                            !is.na(winner) ~ as.character(winner)),
         death_rate_per_100k = (deaths / population) * 100000,
         infection_rate_per_100k = (cases / population) * 100000,
         death_rate = (deaths / population),
         infection_rate = (cases / population))



master_covid_election_with_dates <- master_covid_election_with_dates %>%
  mutate(winner = case_when(is.na(winner) ~ state_win,
                            !is.na(winner) ~ as.character(winner)))

#replacing NAs in other columns with 0's
master_covid_election[is.na(master_covid_election)] <- 0
master_covid_election_with_dates[is.na(master_covid_election_with_dates)] <- 0

#adding state population to covid_master
state_population <- state_population %>%
  rename(state = NAME,
         state_population = value) %>%
  mutate(state = str_to_lower(state)) %>%
  select(-GEOID, -variable)

master_covid_election_with_dates <- master_covid_election_with_dates %>%
  left_join(state_population, by = "state") %>%
  mutate(state_abb = factor(state_abb),
         cases_biden = cases * percent_biden, #adds cases by proportion biden col
         cases_trump = cases * (1 - percent_biden)) #adds cases by proportion trump col

master_covid_election <- master_covid_election %>%
  left_join(state_population, by = "state") %>%
  mutate(state_abb = factor(state_abb),
         cases_biden = cases * percent_biden, #adds cases by proportion biden col
         cases_trump = cases * (1 - percent_biden)) #adds cases by proportion trump col


```

Row {data-height=600}
-------------------------------------

### US Map, Infection & Death Rate with Election Overlay
```{r include=FALSE}
source("us_map.R")
```

```{r}
us_map
```


Plots
=====================================  
    
-------------------------------------


Row {data-height=400}
-------------------------------------
### County Infection Rate by 2020 Presidential Election Winner
 
```{r chart2 cleaning, include=FALSE}
source("dygraph_infectionrate.R")
```

```{r}
dy_infect
```   
    
### Mask-wearing adherence for the three highest COVID-19 cases by state and its counties

*Circle size represents COVID-19 cases per 100K people*
```{r mask, include= FALSE}
source ("mask_final.R")
```

```{r}
mask_plot
```


Dashboard Descriptions
=====================================  
    
-------------------------------------
    
### US Map
    
This map shows all the counties in the US colored by which ever presidential candidate received the majority of votes within the county, and shaded by either the COVID-19 infection rate or COVID-19 death rate for each county. Therefore, dark red counties indicate a Trump victory where there is also a high infection/death rate. Dark blue counties indicate a Biden victory with a high infection/death rate. Light Red for a Trump victory and low infection/death rate, and light blue for a Biden victory with a low infection/death rate.
This map is interactive, so you can pan around and zoom in to find specific counties, toggle between showing the infection and death rate, and also turn the election results on and off.

We got our data from the [NY Times](https://www.nytimes.com/interactive/2020/us/coronavirus-us-cases.html)
   
### County Infection Rate by 2020 Presidential Election Winner

The COVID-19 infection rate per 100k people was calculated for each county in the United States. Dotted lines represent the 3 counties with the highest infection rates that voted either for Biden (blue) or Trump (red) in the 2020 Presidential Election. Solid lines are the average or county infection rates per 100k people colored by voter preference.

### Chart 3
    
Some comment

